---
title: The relationship between unemployment and crime in the Netherlands
author: Romeo Schoonderbeek, 2868407
        Paul Claessens, 2859603
        Furkan Kadir Öztürk, 2852640 
Lecturer : Jack Flitzgerald, Chantal Schouwenaar
        
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("cbsodataR")
install.packages("sf")
```

```{r packages}
require(tidyverse)
require(cbsodataR)
require(sf)
```

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

Our Social problem is to look at the relation between crime and unemployment, since these are often believed to be related with each other. Unemployment can cause financial stress and social exclusion, these factors can lead to criminal activity. Criminal records can also lower the chances of being accepted for a job, which makes criminal activity an easier option. Why this is a relevant issue, even though crime rates in the Netherlands have seen a steady decline. Financial and online crime are relatively growing (Nieuws Sociaal en Groen, 2023). Whilst youth unemployment remains a prominent issue in vulnerable communities. Understanding the relationship between unemployment and crime is highly important for the design of social and economic policies. A study by Verbruggen(2014), reviewed by Fischer(2015) shows a critical perspective on the relationship between crime and unemployment. It shows that employment is generally associated with lower crime rates, however it also shows that unemployment doesn’t necessarily lead to more crime. Whilst Verbruggen's study gives a good insight into the vulnerable Dutch population, broader analyses of this problem are still very limited. How structural shifts affect this relationship for example. Additionally, the understanding of spatial and subgroup variations in the Netherlands are also limited. We aim to contribute towards the limited understanding of the special and subgroup variations.

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data}
# provincelevel
DS_Panel_Unemployment <- read.csv("data/raw_data/unemployment_data_transform.csv")
DS_Panel_Crime <- read.csv("data/raw_data/panel_data_crime.csv")
DS_Population <- read.csv("data/population_data.csv")
area_data_working <- read.csv("data/raw_data/area_data.csv")

# aggragate

agg_unemployment <- read.csv("data/raw_data/aggragate_total_unemployment.csv")
agg_crime <- read.csv("data/raw_data/aggragate_total_crime.csv")
agg_population <- read.csv("data/raw_data/population_data_raw.csv")

# subpopulation

criminaliteit_per_leeftijd <- read.csv("data/raw_data/criminaliteit_per_leeftijd.csv") 

unemployment_age <- read.csv("data/raw_data/unemployment_per_age.csv") 
```

## 2.2 Provide a short summary of the dataset's

The data sets are categorized within 3 categories: province level, aggregate and sub population.

Province level: Includes 4 data sets. Two are self explanatory. Population data gives the population per year per province. Area data gives the area per province (per year). Our main data sets are the employment and crime data sets. DS_panel_Unemployment is raw panel data extracted from CBS. For every year and province the unemployment unemployment rate - (non-employed working population/working population)\*100 -, is given. Similarly, DS_Panel_Crime gives the total amount of reported crimes per year per province. All reported crimes are considered, such as violent, financial and petty crimes. After data cleaning, these provincial panel data sets will be useful to visualize spatial variation.

Aggregate level: These raw data sets focus on the aggregate level of crime and unemployment in the Netherlands. For agg_unemployment the aggregate unemployment rate per year in the Netherlands is given. For agg_crime the same is done, but then with total reported crimes. Each years population is included in the agg_population data set. After data cleaning, these aggregate data sets will be useful to visualize temporal variation, and the aggregate connection between crime and unemployment.

Subpopulation: Similair to previous data sets, the subpopulation data sets provide information of crime and unemployment per age group.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

CBS datasets already come in panel data form. Data cleaning will include removing unnecessary rows and columns, adjusting data frames and merging data sets. 3 main data sets are used. Province panel data, aggregate panel data and data sets used for sub population analysis.

```{r}
# province panel data cleaning

merged_data <- cbind(DS_Panel_Unemployment, DS_Panel_Crime [, 5])

merged_data <-cbind(merged_data, DS_Population [, 4])

merged_data <- merged_data [, -c(1)] 
colnames(merged_data) <- c("Year", "province", "Unemployment_Rate", "total_Crimes", "Population")

merged_data <- mutate(merged_data, crimes_per_capita = (total_Crimes/Population)*100)

write.csv(merged_data, "data/working_data/merged_panel_data_final.csv")

area_data_working <- read.csv("data/raw_data/area_data.csv")

colnames(area_data_working) = c("X", "province", "Area(KM2)")

area_data_working <- area_data_working[6:17 ,]

area_data_working[[3]]  = as.numeric(area_data_working[[3]])

area_data_working[, 3] = area_data_working[, 3]/ 100

flevoland <- data.frame(name = "Flevoland", value = 1417 )
colnames(flevoland) = c("province", "Area(KM2)")

area_data_working <- bind_rows(
  area_data_working[1:5, ], flevoland,  
  area_data_working[6:nrow(area_data_working), ]
)

area_data_working <- area_data_working %>%
  slice(rep(1:n(), each = 12))

merged_panel_data_final <- cbind(merged_data, area_data_working[, 3])

colnames(merged_panel_data_final) <- c("year", "province", "unemployment_rate", "total_crimes", "population", "crimes_per_capita", "area_km2")



# aggregate data cleaning

agg_unemployment <- agg_unemployment[-c(51,52) ,]
agg_unemployment <- agg_unemployment[, -1]

agg_crime <- agg_crime[-82 , ]
agg_crime <- agg_crime[-c(1:31) ,]
agg_crime <- agg_crime[, -1]

colnames(agg_unemployment) = c("year", "unemployment")
colnames(agg_crime) = c("year", "total_crimes")

agg_merged <- cbind(agg_unemployment, agg_crime[, 2])

colnames(agg_merged) = c("year", "unemployment", "total_crime")

agg_merged <- agg_merged %>%
  mutate(total_crime = as.numeric(total_crime)) %>%
  mutate(year = as.numeric(year))

str(agg_merged)

agg_population <-agg_population[, -4]
agg_population <- agg_population[, -1]
agg_population <- agg_population[-c(1:75) ,]

agg_merged <- cbind(agg_merged, agg_population[, 2])

colnames(agg_merged) = c("year", "unemployment", "total_crime", "population")

str(agg_merged)



# subpopulation

crime2021 <- criminaliteit_per_leeftijd %>% filter(Perioden == "2021")

crime2021 <- crime2021 %>% filter(Geboorteland == "Totaal")

crime2021 <- crime2021[, -c(1:2, 4:8)]

colnames(crime2021) <- c("Age", "Crime")

crime2021$Crime = (crime2021$Crime/10000)*100

unemployment_age <- read.csv("data/raw_data/unemployment_per_age.csv")

unemployment_age <- unemployment_age[-c(1:6, 19) ,]

unemployment_age <- unemployment_age[, -1]

colnames(unemployment_age) = c("Age", "Unemployment")

unemployment_age$Unemployment <- gsub(",", ".", unemployment_age$Unemployment)

unemployment_age$Unemployment <- as.numeric(unemployment_age$Unemployment)







```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1 and 2 (Crimes per capita and population density)

```{r gen_var1 and 2}
agg_merged <- agg_merged %>%
  mutate(crime_per_capita = (total_crime/population)*100)

merged_panel_data_final <- mutate(merged_panel_data_final, poplation_density = population/area_km2)



```

Variable 3 Age categories

```{r gen_var3}

crime2021$Age <- ifelse(crime2021$Age %in% c("12 tot 18 jaar", "18 tot 23 jaar"),
                                     "Youth (12–23)", "Adult+ (23+)")

unemployment_age$Age <- ifelse(unemployment_age$Age %in% c("15 tot 20 jaar", "20 tot 25 jaar"),
                             "Youth (15-25)", "Adult (25+)")
```

## 3.3 Visualize temporal variation

```{r}
pivot_agg_data <- agg_merged %>%
  pivot_longer(cols = c(unemployment, crime_per_capita), names_to = "variable", values_to = "value")

ggplot(pivot_agg_data, aes(x = year, y = value, colour = variable )) +
  geom_line() +
  labs(title = "aggragate crime and unemployment through time", x = "year", y = "variable") +
  scale_x_continuous(breaks = seq(1975, 2024, by = 5), limits = c(1972, 2026))
```

## 3.4 Visualize spatial variation

```{r, echo=TRUE, message=FALSE, warning=FALSE}
cbs_maps <- cbs_get_maps()
# the layout of the data.frame is:
str(cbs_maps)
#read the GeoJSON straight from PDOK 
nl_prov <- st_read(
  "https://cartomap.github.io/nl/wgs84/provincie_2014.geojson",
  quiet = TRUE
)
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#Reading the CSV  →  2023 only
crime <- merged_panel_data_final |>
  filter(year == 2023) |>
  mutate(
    province = sub(" \\(PV\\)$", "", province),        
    province = recode(province, "Fryslân" = "Friesland")
  ) |>
  filter(!province %in% "Nederland")
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Province shapes (CBS 2014)
prov_shapes <- read_sf("https://cartomap.github.io/nl/wgs84/provincie_2014.geojson")
# Join data → map
nl_crime <- prov_shapes |>
  left_join(crime, by = c("statnaam" = "province"))
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(nl_crime) +
  geom_sf(aes(fill = crimes_per_capita), color = "white", size = 0.2) +
  scale_fill_gradient(
    low  = "#fee5d9",  # pale red
    high = "#a50f15",  # dark red
    na.value = "grey90",
    name = "Crime\nper capita"
  ) +
  labs(
    title = "Crime per capita by Dutch province – 2023"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

```{r pop-heatmap, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(nl_crime) +
  geom_sf(aes(fill = poplation_density), color = "white", size = 0.2) +
  scale_fill_gradient(
    low  = "#deebf7",  # pale blue
    high = "#08306b",  # dark blue
    na.value = "grey90",
    name = "Population\ndensity"
  ) +
  labs(
    title = "Population density by Dutch province – 2023"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

```{r unemp-heatmap, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(nl_crime) +
  geom_sf(aes(fill = unemployment_rate), color = "white", size = 0.2) +
  scale_fill_gradient(
    low  = "#e5f5e0",  # pale green
    high = "#00441b",  # dark green
    na.value = "grey90",
    name = "Unemployment\nrate"
  ) +
  labs(
    title = "Unemployment rate by Dutch province – 2023"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

<<<<<<< HEAD
=======

>>>>>>> d22ac6077aa71c465abe179b2f0ae4d7a514a3ae
## 3.5 Visualize sub-population variation

```{r visualise_subpopulations}
ggplot(crime2021, aes(x = Age, y = Crime)) +
  geom_boxplot(fill = "yellow", color = "black") +
  labs(
    title = "Boxplot of Crime Suspect Rate by Age Group",
    x = "Age category",
    y = "Crime Rate"
  ) 
ggplot(unemployment_age, aes(x = Age, y = Unemployment)) +
  geom_boxplot(fill = "yellow", color = "black") +
  labs(
    title = "Boxplot of Crime Suspect Rate by Age Group",
    x = "Age category",
    y = "unemployment"
  ) 
```

In the boxplots above, it is clearly visible that unemployment and crime are linked. The Youth category is more vulnarable to unemployment than the adult category (Yeung & Yang, 2020), which in turn correlates with higher crime rates. Kessler et al. (2021) finds that reducing youth unemployment reduces youth crimes.

## 3.6 Event analysis

```{r analysis}
pivot_agg_data <- agg_merged %>%
  pivot_longer(cols = c(unemployment, crime_per_capita), names_to = "variable", values_to = "value")

ggplot(pivot_agg_data, aes(x = year, y = value, colour = variable )) +
  geom_line() +
  labs(title = "aggragate crime and unemployment through time", x = "year", y = "variable") +
  scale_x_continuous(breaks = seq(1975, 2024, by = 5), limits = c(1972, 2026)) +
  geom_vline(xintercept = c(1979, 2008), 
             linetype = "solid", 
             color = "black", 
             size = 0.6) + 
  annotate("text", x = 1974, y = 9, label = "1979\noil-crisis") +
  annotate("text", x = 2017, y = 9, label = "2008\ngreat-recession")
```

Both 1979 oil-crises and the 2008 recession see large spikes in unemployment, yet only 1979 sees an increase in crimp per capita as well. A possible explanation for the movement of crime per capita in accordance with unemployment in 1980 to an inverse movement in 2008 could be the fact that the population is much younger in 1980 compared to 2008 and thus more people are in their peak-crime committing age 18-35 (Bezuinigen En Hervormingen in De Jaren ’80, 2022). Another explanation is that the welfare safety net was much more robust in 2008 than in the 1980’s which means that people who were unemployed where less inclined to do crime (Bezuinigen En Hervormingen in De Jaren ’80, 2022). This shows that while unemployment may be linked in some cases, other factors play a significant role as well.

# Part 4 - Discussion

## 4.1 Discuss your findings

Under each plot certain trends and implacations where already discussed. By reviewing each plot as part of the whole, certain implacations and conclusions can be drawn.

Unemployment is likely a catalyst for crime. I.E. unemployment is related to various social factors which might directly impact potency to commit crimes, such as..

These results can be of interest for policy makers, as not only battling unemployment helps prevent crime, but setting up other social programs might reduce blowbacks of unemployment. Allocating funds towards vulnabrae groups and areas with a high population density will achieve the greatest results.

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: https://github.com/Autelius/Crime-and-Unemployment-Group-Project-FR-GIT.git

## 5.2 Reference list

Use APA referencing throughout your document.
