---
title: The relationship between unemployment and crime in the Netherlands
author: Romeo Schoonderbeek, 2868407
        Paul Claessens, 2859603
        Furkan Kadir Öztürk, 2852640 
Lecturer : Jack Flitzgerald, Chantal Schouwenaar
        
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("cbsodataR")
install.packages("sf")
```

```{r packages}
require(tidyverse)
require(cbsodataR)
require(sf)
```

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Our Social problem is to look at the relation between crime and unemployment, since these are often believed to be related with each other. Unemployment can cause financial stress and social exclusion, these factors can lead to criminal activity. Criminal records can also lower the chances of being accepted for a job, which makes criminal activity an easier option.

Why this is a relevant issue, even though crime rates in the Netherlands have seen a steady decline. Financial and online crime are relatively growing (Nieuws Sociaal en Groen, 2023). Whilst youth unemployment remains a prominent issue in vulnerable communities. Understanding the relationship between unemployment and crime is highly important for the design of social and economic policies.

A study by Verbruggen(2014), reviewed by Fischer(2015) shows a critical perspective on the relationship between crime and unemployment. It shows that employment is generally associated with lower crime rates, however it also shows that unemployment doesn’t necessarily lead to more crime.

Whilst Verbruggen's study gives a good insight into the vulnerable Dutch population, broader analyses of this problem are still very limited. How structural shifts affect this relationship for example. Additionally, the understanding of spatial and subgroup variations in the Netherlands are also limited.

We aim to contribute towards the limited understanding of the special and subgroup variations.

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data}
# provincelevel
DS_Panel_Unemployment <- read.csv("data/raw_data/unemployment_data_transform.csv")
DS_Panel_Crime <- read.csv("data/raw_data/panel_data_crime.csv")
DS_Population <- read.csv("data/population_data.csv")
area_data_working <- read.csv("data/raw_data/area_data.csv")

# aggragate

agg_unemployment <- read.csv("data/raw_data/aggragate_total_unemployment.csv")
agg_crime <- read.csv("data/raw_data/aggragate_total_crime.csv")
agg_population <- read.csv("data/raw_data/population_data_raw.csv")

# subpopulation

criminaliteit_per_leeftijd <- read.csv("data/raw_data/criminaliteit_per_leeftijd.csv") 

unemployment_age <- read.csv("data/raw_data/unemployment_per_age.csv") 



```

## 2.2 Provide a short summary of the dataset(s)

```{r}

```

``` r
```

The data sets are categorized within 3 categories: province level, aggregate and sub population.

Province level: Includes 4 data sets. Two are self explanatory. Population data gives the population per year per province. Area data gives the area per province (per year). Our main data sets are the employment and crime data sets. DS_panel_Unemployment is raw panel data extracted from CBS. For every year and province the unemployment unemployment rate - (non-employed working population/working population)\*100 -, is given. Similarly, DS_Panel_Crime gives the total amount of reported crimes per year per province. All reported crimes are considered, such as violent, financial and petty crimes. After data cleaning, these provincial panel data sets will be useful to visualize spatial variation.

Aggregate level: These raw data sets focus on the aggregate level of crime and unemployment in the Netherlands. For agg_unemployment the aggregate unemployment rate per year in the Netherlands is given. For agg_crime the same is done, but then with total reported crimes. Each years population is included in the agg_population data set. After data cleaning, these aggregate data sets will be useful to visualize temporal variation, and the aggregate connection between crime and unemployment.

Subpopulation: Similair to previous data sets, the subpopulation data sets provide information of crime and unemployment per age group.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

CBS datasets already come in panel data form. Data cleaning will include removing unnecessary rows and columns, adjusting data frames and merging data sets. 3 main data sets are used. Province panel data, aggregate panel data and data sets used for sub population analysis.

```{r}
# province panel data cleaning

merged_data <- cbind(DS_Panel_Unemployment, DS_Panel_Crime [, 5])

merged_data <-cbind(merged_data, DS_Population [, 4])

merged_data <- merged_data [, -c(1)] 
colnames(merged_data) <- c("Year", "province", "Unemployment_Rate", "total_Crimes", "Population")

merged_data <- mutate(merged_data, crimes_per_capita = (total_Crimes/Population)*100)

write.csv(merged_data, "data/working_data/merged_panel_data_final.csv")

area_data_working <- read.csv("data/raw_data/area_data.csv")

colnames(area_data_working) = c("X", "province", "Area(KM2)")

area_data_working <- area_data_working[6:17 ,]

area_data_working[[3]]  = as.numeric(area_data_working[[3]])

area_data_working[, 3] = area_data_working[, 3]/ 100

flevoland <- data.frame(name = "Flevoland", value = 1417 )
colnames(flevoland) = c("province", "Area(KM2)")

area_data_working <- bind_rows(
  area_data_working[1:5, ], flevoland,  
  area_data_working[6:nrow(area_data_working), ]
)

area_data_working <- area_data_working %>%
  slice(rep(1:n(), each = 12))

merged_panel_data_final <- cbind(merged_data, area_data_working[, 3])

colnames(merged_panel_data_final) <- c("year", "province", "unemployment_rate", "total_crimes", "population", "crimes_per_capita", "area_km2")



# aggregate data cleaning

agg_unemployment <- agg_unemployment[-c(51,52) ,]
agg_unemployment <- agg_unemployment[, -1]

agg_crime <- agg_crime[-82 , ]
agg_crime <- agg_crime[-c(1:31) ,]
agg_crime <- agg_crime[, -1]

colnames(agg_unemployment) = c("year", "unemployment")
colnames(agg_crime) = c("year", "total_crimes")

agg_merged <- cbind(agg_unemployment, agg_crime[, 2])

colnames(agg_merged) = c("year", "unemployment", "total_crime")

agg_merged <- agg_merged %>%
  mutate(total_crime = as.numeric(total_crime)) %>%
  mutate(year = as.numeric(year))

str(agg_merged)

agg_population <-agg_population[, -4]
agg_population <- agg_population[, -1]
agg_population <- agg_population[-c(1:75) ,]

agg_merged <- cbind(agg_merged, agg_population[, 2])

colnames(agg_merged) = c("year", "unemployment", "total_crime", "population")

str(agg_merged)



# subpopulation

crime2021 <- criminaliteit_per_leeftijd %>% filter(Perioden == "2021")

crime2021 <- crime2021 %>% filter(Geboorteland == "Totaal")

crime2021 <- crime2021[, -c(1:2, 4:8)]

colnames(crime2021) <- c("Age", "Crime")

crime2021$Crime = (crime2021$Crime/10000)*100

unemployment_age <- read.csv("data/raw_data/unemployment_per_age.csv")

unemployment_age <- unemployment_age[-c(1:6, 19) ,]

unemployment_age <- unemployment_age[, -1]

colnames(unemployment_age) = c("Age", "Unemployment")

unemployment_age$Unemployment <- gsub(",", ".", unemployment_age$Unemployment)

unemployment_age$Unemployment <- as.numeric(unemployment_age$Unemployment)







```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1 and 2 (Crimes per capita and Unemployment per capita)

```{r gen_var1 and 2}
agg_merged <- agg_merged %>%
  mutate(crime_per_capita = (total_crime/population)*100)

merged_panel_data_final <- mutate(merged_panel_data_final, poplation_density = population/area_km2)

```

Variable 3 Age categories

```{r gen_var3}

crime2021$Age <- ifelse(crime2021$Age %in% c("12 tot 18 jaar", "18 tot 23 jaar"),
                                     "Youth (12–23)", "Adult+ (23+)")

unemployment_age$Age <- ifelse(unemployment_age$Age %in% c("15 tot 20 jaar", "20 tot 25 jaar"),
                             "Youth (15-25)", "Adult (25+)")
```

## 3.3 Visualize temporal variation

```{r}
pivot_agg_data <- agg_merged %>%
  pivot_longer(cols = c(unemployment, crime_per_capita), names_to = "variable", values_to = "value")

ggplot(pivot_agg_data, aes(x = year, y = value, colour = variable )) +
  geom_line() +
  labs(title = "aggragate crime and unemployment through time", x = "year", y = "variable") +
  scale_x_continuous(breaks = seq(1975, 2024, by = 5), limits = c(1972, 2026))
```

## 3.4 Visualize spatial variation

```{r, echo=TRUE, message=FALSE, warning=FALSE}
cbs_maps <- cbs_get_maps()
# the layout of the data.frame is:
str(cbs_maps)
#read the GeoJSON straight from PDOK 
nl_prov <- st_read(
  "https://cartomap.github.io/nl/wgs84/provincie_2014.geojson",
  quiet = TRUE
)
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#Reading the CSV  →  2023 only
crime <- merged_panel_data_final |>
  filter(year == 2023) |>
  mutate(
    province = sub(" \\(PV\\)$", "", province),        
    province = recode(province, "Fryslân" = "Friesland")
  ) |>
  filter(!province %in% "Nederland")
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Province shapes (CBS 2014)
prov_shapes <- read_sf("https://cartomap.github.io/nl/wgs84/provincie_2014.geojson")
# Join data → map
nl_crime <- prov_shapes |>
  left_join(crime, by = c("statnaam" = "province"))
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(nl_crime) +
  geom_sf(aes(fill = crimes_per_capita), color = "white", size = 0.2) +
  scale_fill_gradient(
    low  = "#fee5d9",  # pale red
    high = "#a50f15",  # dark red
    na.value = "grey90",
    name = "Crime\nper capita"
  ) +
  labs(
    title = "Crime per capita by Dutch province – 2023"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

```{r pop-heatmap, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(nl_crime) +
  geom_sf(aes(fill = poplation_density), color = "white", size = 0.2) +
  scale_fill_gradient(
    low  = "#deebf7",  # pale blue
    high = "#08306b",  # dark blue
    na.value = "grey90",
    name = "Population\ndensity"
  ) +
  labs(
    title = "Population density by Dutch province – 2023"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

```{r unemp-heatmap, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(nl_crime) +
  geom_sf(aes(fill = unemployment_rate), color = "white", size = 0.2) +
  scale_fill_gradient(
    low  = "#e5f5e0",  # pale green
    high = "#00441b",  # dark green
    na.value = "grey90",
    name = "Unemployment\nrate"
  ) +
  labs(
    title = "Unemployment rate by Dutch province – 2023"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
ggplot(crime2021, aes(x = Age, y = Crime)) +
  geom_boxplot(fill = "yellow", color = "black") +
  labs(
    title = "Boxplot of Crime Suspect Rate by Age Group",
    x = "Age category",
    y = "Crime Rate"
  ) 
ggplot(unemployment_age, aes(x = Age, y = Unemployment)) +
  geom_boxplot(fill = "yellow", color = "black") +
  labs(
    title = "Boxplot of Crime Suspect Rate by Age Group",
    x = "Age category",
    y = "unemployment"
  ) 
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
pivot_agg_data <- agg_merged %>%
  pivot_longer(cols = c(unemployment, crime_per_capita), names_to = "variable", values_to = "value")

ggplot(pivot_agg_data, aes(x = year, y = value, colour = variable )) +
  geom_line() +
  labs(title = "aggragate crime and unemployment through time", x = "year", y = "variable") +
  scale_x_continuous(breaks = seq(1975, 2024, by = 5), limits = c(1972, 2026)) +
  geom_vline(xintercept = c(1979, 2008), 
             linetype = "solid", 
             color = "black", 
             size = 0.6) + 
  annotate("text", x = 1974, y = 9, label = "1979\noil-crisis") +
  annotate("text", x = 2017, y = 9, label = "2008\ngreat-recession")
```

Graph of Aggregate Crime per Capita and Aggregate unemployment in the Netherlands through time. Furthermore our temporal events: the Oil crisis 1979 and the Financial crisis 2008 are highlighted. 

# Part 4 - Discussion

After the Oil-crisis in 1979, The Netherlands would experience significant unemployment and crime per capita. According to Ingeschiedenis.nl this was due to The Netherlands having a high budget deficit and to little government intervention in accordance with the Keynesian economic view of the time period (Bezuinigen En Hervormingen in De Jaren ’80, 2022). After the 2008 financial crisis The Netherlands would find itself in a similar position with unemployment rising, however crime per capita would fall in the next 5 years to historic lows. What explains this phenomenon? An explanation by Ingschiedenis.nl is that while the Netherlands decided after 1985 to reduce wages and the budget deficit, The Netherlands in 2008 decided to increase spending, this decision would be later reversed in the 2010’s to a policy of fiscal conservatism (Bezuinigen En Hervormingen in De Jaren ’80, 2022). A possible explanation for the movement of crime per capita in accordance with unemployment in 1980 to an inverse movement in 2008 could be the fact that the population is much younger in 1980 compared to 2008 and thus more people are in their peak-crime committing age 18-35 (Bezuinigen En Hervormingen in De Jaren ’80, 2022). Another explanation is that the welfare safety net was much more robust in 2008 than in the 1980’s which means that people who were unemployed where less inclined to do crime (Bezuinigen En Hervormingen in De Jaren ’80, 2022).


## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Problem Motivation:
Fischer, T. (2015). De relatie tussen werk en criminaliteit uitgeplozen. Tijdschrift voor Criminologie, 57(3), 334–338. https://www.boomportaal.nl/tijdschrift/TvC/TvC_0165-182X_2015_057_003_008 
Nieuws Sociaal en Groen. (2023). Criminaliteit: de stand van zaken in Nederland. Retrieved from https://nieuwsociaalengroen.nl/criminaliteit/
Verbruggen, J. (2014). Previously institutionalized youths on the road to adulthood: A longitudinal study on employment and crime. [PhD-Thesis - Research and graduation internal, Vrije Universiteit Amsterdam].
Data:
Statline. (2025). Arbeidsdeelname; provincie, 2013–2024 [Data set]. Centraal Bureau voor de Statistiek. https://opendata.cbs.nl/#/CBS/nl/dataset/85268NED/table?dl=BE1FB 
Cbs. (2025). Geregistreerde criminaliteit; soort misdrijf, regio, 2010-2024 [Data set]. Centraal Bureau voor de Statistiek. https://opendata.cbs.nl/#/CBS/nl/dataset/85268NED/table?dl=BE1FB 
Statline. (2025). Bevolkingsontwikkeling; regio per maand, 2002-2024 [Data set]. Centraal Bureau voor de Statistiek. https://opendata.cbs.nl/#/CBS/nl/dataset/37230ned/table 
Statline. (2006). Volkstelling; oppervlakten, 1930 [Data set]. Centraal Bureau voor de Statistiek. https://opendata.cbs.nl/#/CBS/nl/dataset/71118NED/table 
Statline. (2025). Geregistreerde criminaliteit, 1948-2024 [Data set]. Centraal Bureau voor de Statistiek. https://opendata.cbs.nl/#/CBS/nl/dataset/83723NED/table 
Statline. (2025). Verdachten; geslacht, leeftijd, herkomst, opleiding, huishoudensinkomen, 2010-2024 [Data set]. Centraal Bureau voor de Statistiek. https://opendata.cbs.nl/#/CBS/nl/dataset/85658NED/table   
Statline. (2025). Arbeidsdeelname; kerncijfers, 2013-2024 [Data set]. Centraal Bureau voor de Statistiek. https://opendata.cbs.nl/#/CBS/nl/dataset/85264NED/table?dl=BE1F8 
Descriptions:
Bezuinigen en hervormingen in de jaren ’80. (2022, June 21).  IsGeschiedenis. https://isgeschiedenis.nl/nieuws/bezuinigen-en-hervormingen-in-de-jaren-80
