---
title: The relationship between unemployment and crime in the Netherlands
author: Romeo Schoonderbeek, 2868407
        Paul Claessens, 2859603
Lecturer : Jack Flitzgerald, Chantal Schouwenaar
        
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
```

```{r packages}
require(tidyverse)
```

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data}
DS_Panel_Unemployment <- read.csv("data/working_data/panel_data_unemployement.csv")
DS_Panel_Crime <- read.csv("data/working_data/panel_data_crime.csv")
DS_Population <- read.csv("data/population_data.csv")
area_data_working <- read.csv("data/raw_data/area_data.csv")

###### not raw data, isert raw data from cvs (possible that some additional cleaning should be done)

agg_unemployment <- read.csv("data/raw_data/aggragate_total_unemployment.csv")
agg_crime <- read.csv("data/raw_data/aggragate_total_crime.csv")
agg_population <- read.csv("data/raw_data/population_data_raw.csv")
```
## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

CBS datasets already come in panel data form. Data cleaning will include removing unncessary rows and columns, adjusting data frames and merging data sets. 3 main data sets are used. Province panel data, aggregate panel data and data sets used for sub population analysis. 

```{r}
# province panel data cleaning

merged_data <- cbind(DS_Panel_Unemployment, DS_Panel_Crime [, 5])

merged_data <-cbind(merged_data, DS_Population [, 4])

merged_data <- merged_data [, -c(1:2)] 
colnames(merged_data) <- c("Year", "province", "Unemployment_Rate", "total_Crimes", "Population")

merged_data <- mutate(merged_data, crimes_per_capita = (total_Crimes/Population)*100)

write.csv(merged_data, "data/working_data/merged_panel_data_final.csv")

area_data_working <- read.csv("data/raw_data/area_data.csv")

colnames(area_data_working) = c("X", "province", "Area(KM2)")

area_data_working <- area_data_working[6:17 ,]

area_data_working[[3]]  = as.numeric(area_data_working[[3]])

area_data_working[, 3] = area_data_working[, 3]/ 100

flevoland <- data.frame(name = "Flevoland", value = 1417 )
colnames(flevoland) = c("province", "Area(KM2)")

area_data_working <- bind_rows(
  area_data_working[1:5, ], flevoland,  
  area_data_working[6:nrow(area_data_working), ]
)

area_data_working <- area_data_working %>%
  slice(rep(1:n(), each = 12))

merged_panel_data_final <- cbind(merged_data, area_data_working[, 3])

colnames(merged_panel_data_final) <- c("year", "province", "unemployment_rate", "total_crimes", "population", "crimes_per_capita", "area_km2")

merged_panel_data_final <- mutate(merged_panel_data_final, poplation_density = population/area_km2)

# aggregate data cleaning

```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r crime-heatmap, echo=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(ggplot2)
library(cbsodataR)
library(dplyr)       
library(readr)
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
cbs_maps <- cbs_get_maps()
# the layout of the data.frame is:
str(cbs_maps)


#read the GeoJSON straight from PDOK 
nl_prov <- st_read(
  "https://cartomap.github.io/nl/wgs84/provincie_2014.geojson",
  quiet = TRUE
)
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#Reading the CSV  →  2023 only
crime <- read_csv("data/working_data/merged_panel_data_final.csv",
                  show_col_types = FALSE) |>
  filter(year == 2023) |>
  mutate(
    province = sub(" \\(PV\\)$", "", province),        
    province = recode(province, "Fryslân" = "Friesland")
  ) |>
  filter(!province %in% "Nederland")
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Province shapes (CBS 2014)
prov_shapes <- read_sf("https://cartomap.github.io/nl/wgs84/provincie_2014.geojson")


# Join data → map
nl_crime <- prov_shapes |>
  left_join(crime, by = c("statnaam" = "province"))
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(nl_crime) +
  geom_sf(aes(fill = crimes_per_capita), color = "white", size = 0.2) +
  scale_fill_gradient(
    low  = "#fee5d9",  # pale red
    high = "#a50f15",  # dark red
    na.value = "grey90",
    name = "Crime\nper capita"
  ) +
  labs(
    title = "Crime per capita by Dutch province – 2023"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

```{r pop-heatmap, echo=TRUE, message=FALSE, warning=FALSE}

ggplot(nl_crime) +
  geom_sf(aes(fill = poplation_density), color = "white", size = 0.2) +
  scale_fill_gradient(
    low  = "#deebf7",  # pale blue
    high = "#08306b",  # dark blue
    na.value = "grey90",
    name = "Population\ndensity"
  ) +
  labs(
    title = "Population density by Dutch province – 2023"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

```{r unemp-heatmap, echo=TRUE, message=FALSE, warning=FALSE}

ggplot(nl_crime) +
  geom_sf(aes(fill = unemployment_rate), color = "white", size = 0.2) +
  scale_fill_gradient(
    low  = "#e5f5e0",  # pale green
    high = "#00441b",  # dark green
    na.value = "grey90",
    name = "Unemployment\nrate"
  ) +
  labs(
    title = "Unemployment rate by Dutch province – 2023"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
